\chapter{Advanced Techniques}\label{ch:advanced-techniques}

\q{What others do may be the stimulus of our feelings, but never the cause.}{Marshall B. Rosenberg}

\q{So many times I fucked up. I talked, when I should have listened. I was harsh, when I should have been tender.}{Ad Astra}

% TODO "you make me feel"; responsible emotional response
% TODO stimulus vs cause. response-able not others emotions, but own behavior. see stoicism difficulty...

% TODO quote stoicism; not events, but how we think about

"we are not responsible for the other person's emotional response"

% use: protective use of force (if needed, for children's safety), but never punishment


\section{Communication Square}\label{sec:communication-square}
% TODO schulz thun

\section{The Work}\label{sec:the-work}

byron katie

\section{Anger}\label{sec:anger}

\subsection{Healthy expression of anger}\label{subsec:healthy-expression-of-anger}

\q{If we wish to express anger fully, the first step is to divorce the other person from any responsibility for our anger.}{Marshall B. Rosenberg}

quote from "basics chapter": I think one of the most difficult aspect of stop blaming, of stop trying to find a scape goat, the one who is guilty, is, that it is even more painful to just sit with the fact that there is no one to blame.
It is just so damn tragic, and there is so much pain, and it can't be redirected to someone else in the form of hate and anger, leading to destructive behavior.
Anger and hate are very powerful emotions, which have a lot of active force and potential for change, and that's why they might be more comfortable for some people to stay with.
But once we acknowledge reality for what it is, we see that most of the time, our anger is a cover up for sadness/pain/grief or fear. (example: traffic; you idiot, instead "i was afraid my health would be at stake")

\q{We are never angry because of what others say or do. It is our thinking that makes us angry.}{Marshall B. Rosenberg}

Osho: vomit on the ground (or bathroom), not on the person.

\q{Resentment is like swallowing poison and expecting the other person to die.}{Bert Ghezzi}

Story with my father....

% TODO celebrate your jackals, project and own it fully. radical honesty, zero censorship. it's not agreeing to that voice, but acknowledging the pain underneath. it's the first step of the process to revelation. otherwise (spiritual) bypassing. life is all, every emotion, happiness not better than suffering, love and hate, we need all, just how we express it. listen to anger, not believing it, but giving it space. toxic positivity tends to love out of fear of hate. go full circle, full range. emotions are not semi permeable. all or nothing. keep pain out, keep also love out. welcome all emotions, like rumis story.

